---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
Load the data frame containing the activity
```{r Loading The Data, cache=TRUE ,echo=TRUE} 
#setwd("c:/wb/repdata/RepData_PeerAssessment1")
if (!file.exists("./activity.csv")) 
{
  f<- download.file(url = 'http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip' , destfile = "activity.zip") 
  activity.filename <-unzip("./activity.zip")
}
dfActivity <- read.csv ( file = "./activity.csv")

```

## What is mean total number of steps taken per day?
For this part of the assignment, you can ignore the missing values in the dataset.
Make a histogram of the total number of steps taken each day
```{r}
library(reshape)
dfActivity.perday <- aggregate( steps~date, data = dfActivity,FUN  =sum ) 
hist(dfActivity.perday$steps , breaks = 50 )
```

Calculate and report the mean and median total number of steps taken per day
```{r}
mean(dfActivity.perday$steps)
median(dfActivity.perday$steps)
```


## What is the average daily activity pattern?
Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
dfActivity.perinterval <- aggregate( steps~interval, data = dfActivity,FUN  =mean)
plot( y=dfActivity.perinterval$steps , x=dfActivity.perinterval$interval , type="l",  col="blue")
```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
maxNumStepsInAInterval       <- max( dfActivity.perinterval$steps )
maxNumStepsInAInterval.index <- match( maxNumStepsInAInterval, dfActivity.perinterval$steps)
tmpi<-dfActivity.perinterval[maxNumStepsInAInterval.index,]
```
The following interval  on average has the higest numbr of steps.
```{r}
tmpi$interval
tmpi$steps
```



## Imputing missing values
Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.
Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r}
totalRowsWithNA <- sum(is.na( dfActivity$steps))
```

Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.
Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}
dfActivity.NAfixed <- dfActivity
msk <- is.na(dfActivity$steps)  #the mask for the rows that have NA result
m1 <- merge(dfActivity[msk, ] , dfActivity.perinterval , books, by.x = "interval", by.y = "interval") #Get the average for the intervals that are NA
dfActivity.NAfixed[msk, ]$steps <- m1$steps.y
```
each interval that contain NA steps, has its NA swaped with the  average steps for that interval


Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. 
```{r}
dfActivity.NAFixed.perday <- aggregate( steps~date, data = dfActivity.NAfixed ,FUN  =sum)
hist(dfActivity.NAFixed.perday$steps , breaks = 50 )
```
Mean and median total number of steps taken per day on the naFixed dataset:
```{r}
mean( dfActivity.NAFixed.perday$steps)
median(dfActivity.NAFixed.perday$steps)
```

```
Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?
```{r}
impact <- merge( dfActivity.NAFixed.perday ,   dfActivity.perday ,  by.x = "date", by.y = "date", all=TRUE)
impact

meanImpact<-mean( dfActivity.NAFixed.perday$steps) - mean(dfActivity.perday$steps)
medianImpact<-median(dfActivity.NAFixed.perday$steps) - median(dfActivity.perday$steps)
meanImpact
medianImpact
```

As it can be seen, there are some days for which no number of steps is recroded. In the dfACtivity.NAFixed, I fixed this issue by using the interval average in the place of NA. No change in the mean as we swaped the NAs with mean, but median has changed. This is naturally  do to inclusion of more means. 


## Are there differences in activity patterns between weekdays and weekends?
For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.
Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
```{r}
mskIsWeekend = weekdays( as.Date( dfActivity.NAfixed$date) ) %in% c("Sunday" , "Saturday" )  
dfActivity.NAfixed$daytype = "weekday"
dfActivity.NAfixed[mskIsWeekend, ]$daytype = "weekend"
dfActivity.NAfixed$daytype = as.factor ( dfActivity.NAfixed$daytype )
str(dfActivity.NAfixed)
```


Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r}
library(lattice)
par(mfrow=c(1,1))
aggSteps <- aggregate ( steps~ interval + daytype ,  data=dfActivity.NAfixed , FUN=mean)

xyplot( steps~interval|daytype, 
        type = "l",
        xlab="Interval", ylab="Average Steps Taken Per Interval", 
        main="Steps taken in each interval in the day, by day type" , 
        layout=c(1, 2),
        data = aggSteps 
        )
```

* At both tails (1 to 500 and 2000 to 2400), weekends and weekdays are similar: little activity during sleep/rest
* At 500 to 1000, there is much greater activity during weekdays
* At 1000 to 1500, there is slightly greater activity during weekdays
* At 1500 to 2000, there is greater activity during weekdays
